---
title: "Data Visualization in R"
author: "Clay Ford, Statistical Research Consultant, UVA Library"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Quick Intro to R Notebooks and R Markdown

This is an R Markdown Notebook. It combines markdown, a plain text formatting syntax, and R code. When you execute R code within the notebook, the output appears beneath the code. 

This file was created in RStudio by going to File...New File...R Notebook.

R code needs to be in "chunks" for it to work. Below is an example of an R code chunk. It makes a parabola using base R graphics (not ggplot2).

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter* (Win/Linux) or *Cmd+Shift+Return* (Mac). 

```{r}
x <- seq(-1, 1, by = 0.01)
y <- x^2
plot(x, y, type = "l")
```

To hide the output, click the Expand/Collapse output button. To clear results (or an error), click the "x". 

You can also press *Ctrl+Enter* (Win/Linux) or *Cmd+Return* (Mac) to run one line of code at a time (instead of the entire chunk).

Add a new R code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac).  

## CODE ALONG 0

Insert a new R code chunk below and type/run the code: 22/7


## Import data

Let's import some data which we'll visualize today using ggplot2. The data we'll work with is Albemarle County real estate data which was downloaded from the Office of Geographic Data Services. We'll use a random sample of the data.

```{r}
URL <- 'https://raw.githubusercontent.com/clayford/dataviz_with_ggplot2/master/alb_homes.csv'
homes <- read.csv(file = URL)
```

Let's look at the first few rows:

```{r}
head(homes)
```

Variable name definitions:
- *totalvalue*: total assessed value of home and property
- *finsqft*: size of house in number square feet
- *lotsize*: size of land on which home is located, in acres
- *bedroom*: number of bedrooms
- *fullbath*: number of full bathrooms (toilet, sink and bath)
- *yearbuilt*: year house was built
- *esdistrict*: the elementary school the home feeds into
- *msdistrict*: the middle school the home feeds into
- *hsdistrict*: the high school the home feeds into
- *censustract*: the census tract the home is located in
- *condition*: assessed condition of home (Substandard, Poor, Fair, Average, Good, Excellent)
- *cooling*: 'Central Air' versus 'No Central Air'
- *age*: of the house in years as of 2025
- *fp*: indicator if house has fireplace (0=no, 1=yes)


## Intro to ggplot2

The ggplot2 package implements *The Grammar of Graphics* as defined in the book of the same name by Leland Wilkinson.

It requires your data be in a data frame.

Let's do a quick example. How does the total value of a home relate to its size in finished square feet? Plot totalvalue versus finsqft to create a scatter plot. 

First we need to load the ggplot2 package. Only do this once per R session.

```{r}
# install.packages('ggplot2')
library(ggplot2)
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point()
```

What just happened?

- The `ggplot()` function creates a plotting area for a data frame
- The `aes()` function maps variables in the data frame to aesthetic properties of geometric shapes
- The `geom_point()` function specifies we want to use points
- combine these functions with `+`

Don't forget the parentheses on `geom_point()`! 

*Most of what we do with ggplot2 takes this general form*

```
ggplot(data) +
  aes() +
  geom_xxxxx()
```

Obviously it can get much more complicated. But for exploratory visualization this often sufficient.

It's worth pointing out that many R users place the `aes()` function in the `ggplot()` function, like this:

```{r}
ggplot(homes, aes(x = finsqft, y = totalvalue)) +
  geom_point()
```



## CODE ALONG 1

Do younger (more recently built) homes have a tendency to be bigger than older homes? Create a scatterplot of finsqft (y-axis) versus age (x-axis). Reminder: Insert a new code chunk by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac). Try entering `alpha = 1/5` in `geom_point()`. What does it do?




## More aesthetics and facets

Points not only have positions in a plane, but can be different sizes and have different shapes and colors. We can map variables in our data frame to these aesthetics. 

How does the relationship between totalvalue and finsqft differ between homes with and without central air?

Plot totalvalue versus finsqft to create a scatter plot, but also *color* the points according to whether or not the homes have central air (cooling). Notice we add `color = cooling` to the `aes()` function below: 

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = cooling) +
  geom_point(alpha = 1/3) 
```

You may not like the default colors. To change the colors, we need to change the color *scale*. The easiest way is to use the `scale_color_manual()` function and list the colors in a vector using the `values` argument. To see available colors, type `colors()` in the console. Can also visit <https://r-charts.com/colors/>.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = cooling) +
  geom_point() +
  scale_color_manual(values = c("blue", "red"))
```

About 4.5% of people are colorblind. To use colorblind-friendly color palettes, we can use the {RColorBrewer} package. To see which palettes are colorblind friendly, use the `display.brewer.all()` function with `colorblindFriendly = TRUE`.

```{r}
library(RColorBrewer)
display.brewer.all(colorblindFriendly = TRUE)
```

To use one of the palettes in {ggplot2}, use the `scale_color_brewer()` function and set the palette using the `palette` argument. For example, here's how we use the 'Set2' palette.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = cooling) +
  geom_point() +
  scale_color_brewer(palette = "Set2")
```

We can also map the *size* and *shape* of points to variables. Below I map cooling to shape and lotsize to size. 

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, shape = cooling, size = lotsize) +
  geom_point() 
```

Notice the overplotting. Perhaps we prefer to see separate scatterplots for each cooling category. We can achieve that with *facets*. The function to use is `facet_wrap()` with the variable that we want to facet on. In this case it's cooling. Notice we precede the variable with a tilde ~. We can read that as "facet by cooling"  

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, size = lotsize) +
  geom_point(alpha = 1/2) +
  facet_wrap(~ cooling)

```

We can have multiple geoms in a plot. Let's say we want to add a *smooth trend line* that summarizes the association between finsqft and totalvalue. We can add `geom_smooth()` to our code:

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point(alpha = 1/2) +
  geom_smooth() +
  facet_wrap(~ cooling) 

```


## CODE ALONG 2

How is totalvalue related to lotsize within each of the three high school districts (hsdistrict)?

Create a scatterplot of totalvalue (y-axis) vs lotsize (x-axis), faceted by hsdistrict. Add smooth trend lines to summarize the relationship.




## Zooming in on data

Albemarle county has homes with very large totalvalue and lotsize values. Including them in our plots means we "zoom out" to accommodate them which means we're "far away" from the majority of homes.  

To zoom in on data with ggplot2, we use the `coord_cartesian()` function and define the limits to view with the `xlim` and `ylim` arguments. Below we zoom in on the y-axis to see homes that range in value from from 0 to $2,000,000 and on the x-axis to see homes that range in finsqft from 0 to 4000. Notice we define that lower and upper limits as a vector using the `c()` function.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point() +
  geom_smooth() + 
  coord_cartesian(xlim = c(0, 4000), ylim = c(0, 2e6))

```


The numbers are a little hard to read on the y-axis. We can format them as dollar amounts using the `labels` argument in the `scale_y_continuous()` function. To do this we need to load the scales package, which is installed when you install the ggplot2 package. 

```{r}
library(scales) # for the dollar function
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(xlim = c(0, 4000), ylim = c(0, 2e6)) + 
  scale_y_continuous(labels = dollar) 

```


## CODE ALONG 3

Modify your code from Exercise 2 to zoom in on the y-axis to see home prices ranging from 0 to $1,000,000, and zoom in on the x-axis to view homes on lots ranging in size from 0 to 5 acres. Also use the dollar function to format the home prices on the y-axis.



## Visualizing counts

Recall the condition variable. We can quickly get counts of each condition using the `table()` function:

```{r}
table(homes$condition)
```

One way to visualize counts is with a bar plot. We can create a bar plot in ggplot2 with `geom_bar()`. Notice it automatically generates the counts for us.

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar()
```

If we want the x-axis ordered according to condition, it's best to define that ordering in the data frame. We can use the `factor` function to achieve this.

```{r}
homes$condition <- factor(homes$condition, 
                          levels = c('Very Poor', 'Poor', 'Fair', 
                                     'Average', 'Average Plus', 
                                     'Good', 'Excellent'))
```

Now when we generate the bar plot, the x-axis in the order as defined in the data frame.

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar()
```


One of the aesthetics of bars is their fill color. We can map a variable from our data frame to the *fill* aesthetic. For example, let's see counts of conditions for homes with and without central air (cooling). Notice we add `fill = cooling` to the `aes()` function.

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar()

```

The default result is to stack the bars. We can set them side-by-side setting the position argument to "dodge".

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar(position = "dodge")

```


The large number of Average homes with central air makes it difficult to see the counts for the other categories. An alternative approach is to view the *proportion* of homes with and without central air within each condition. We can do this by setting the position argument to "fill".

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar(position = "fill") +
  ylab("proportion")

```

The y-axis now represents proportions. 

We can also change the color palette. Notice this time we need to use `scale_fill_brewer` because we're changing the *fill* aesthetic.

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set2") +
  ylab("proportion")
```


## CODE ALONG 4

Given a home's condition, is it more likely to be in a particular high school district? For example, given all the homes in Excellent condition, is there a higher proportion of them in one of the three high school districts?

Create a bar plot of condition (condition on x-axis), filled by hsdistrict. Set `position = 'fill'` so we can visualize within each condition the proportion belonging to a high school district.




## Plotting continuous values vs groups

How does the totalvalue of homes differ between elementary school districts? One way to visualize this is with a boxplot. We can do this with `geom_boxplot()`.


```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_boxplot()
```

The x-axis is hard to read. This happens sometimes. One solution is to simply rotate the plot. We can do that with the `coord_flip()` function. The `coord_flip()` function also allows us to zoom in on a plot with the xlim and ylim arguments. Let's zoom in on the y-axis and look at homes less than 2 million in totalvalue.

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_boxplot() +
  coord_flip(ylim = c(0,2e6))

```


- The width of the box (along the y axis) represents the middle 50% of the data. It is called the Interquartile Range (IQR).
- The line in the box is the median. 
- The top of the box is the 75th percentile. 
- The bottom of the box is the 25th percentile. 
- The "whiskers" extending out of the box are to the highest and lowest values not to exceed 1.5 * IQR
- The dots are considered "outliers" relative to the rest of the data within each elementary school district.


A similar plot is the violin plot, which allows us to compare the distributions between categories using smooth density plots. Use `geom_violin()`

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_violin() +
  coord_flip(ylim = c(0,1e6))
```

Boxplots and violin plots obscure the total number within in each group. That's not necessarily a bad thing. Too much information in a plot can be overwhelming. However, ggplot allows us to combine geoms, so we could combine a 1-dimensional scatterplot (sometimes called a "stripchart") with a boxplot or violin plot. The `position_jitter` functions adds a little random noise to data to help spread them out. 

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_point(position = position_jitter(width = 0.2, height = 0)) +
  geom_violin(alpha = 1/2) +
  coord_flip(ylim = c(0,1e6))

```


## CODE ALONG 5

How are the ages of homes distributed between the elementary school districts? Are there some elementary schools that serve mostly newer homes? Create a boxplot of age of home by esdistrict. 




## Line plots over time

How has the median size (finsqft) of houses changed over time (yearbuilt)? It seems like homes have been getting bigger. Can we visualize how the median finsqft has changed over the years?

To do this we need to get the median finsqft per year. The result needs to be a data frame. There are many ways to do this in R. Here is one way.

```{r}
homes_median_sf <- aggregate(finsqft ~ yearbuilt, data = homes, median)
tail(homes_median_sf, n = 10)
```

Let's plot finsqft vs yearbuilt using dots and lines.

```{r}
ggplot(homes_median_sf) +
  aes(x = yearbuilt, y = finsqft) +
  geom_point() +
  geom_line()
```

There's a lot of variability in median home size in finsqft over time! But it tends to settle down around 1950 and then slowly increase. 


How has lotsize changed over time? Let's plot median lotsize by yearbuilt versus yearbuilt. Again we need to make a new data frame that contains yearbuilt and median lotsize per year.

```{r}
homes_median_lotsize <- aggregate(lotsize ~ yearbuilt, data = homes, median)

ggplot(homes_median_lotsize) +
  aes(x = yearbuilt, y = lotsize) +
  geom_line() 
```

After 1950 it looks most new homes are built on small lots.

We can zoom in and get more detail on the last 75 years (since 1950). Notice we use the `labs()` function to add a title and customize the axis labels.

```{r}
ggplot(homes_median_lotsize) +
  aes(x = yearbuilt, y = lotsize) +
  geom_line() +
  coord_cartesian(xlim = c(1950,2025), ylim = c(0,10)) +
  labs(title = "Median Lot Size by Year Built", 
       x = "Year Built", 
       y = "Median Lot Size")

```



## CODE ALONG 6

How has the mean number of fullbaths in a home changed over time (yearbuilt), since 1950?

The following code creates a data frame of mean fullbath by yearbuilt for 1950 to present.

```{r}
homes_mean_fb <- aggregate(fullbath ~ yearbuilt, 
                           data = homes, mean, 
                           subset = yearbuilt >= 1950)
head(homes_mean_fb)
```

Create a line plot showing how the mean number of fullbaths has changed over time (yearbuilt). Try adding a smooth trend line using `geom_smooth()`.



## Two last items

Two more topics worth mentioning:

1. Changing the appearance of plots
2. Saving plots as jpg, png, tif, etc.

The appearance of plots are dictated by themes. Some different themes are included with ggplot2. For example, `theme_minimal()`. Simply tack it on to the end of a plot.

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = lotsize) +
  geom_point(position = position_jitter(width = 0.1, height = 0)) +
  coord_flip(ylim = c(0,50)) +
  theme_minimal()
```

There are whole R packages that provide nothing but more themes for ggplot2. For example, see the {ggthemes} and {ggpubr} packages. Here's an example using `theme_pubr()` from the {ggpubr} package 


```{r}
# install.packages("ggpubr")
library(ggpubr)
ggplot(homes) +
  aes(x = esdistrict, y = lotsize) +
  geom_point(position = position_jitter(width = 0.1, height = 0)) +
  coord_flip(ylim = c(0,50)) +
  theme_pubr()
```


Saving plots can be done with the `ggsave()` function. The file name you provide determines the type of plot saved. By default, it will save the last plot you created as a 7 in. x 7 in. image.

```{r}
ggsave("es_lotsize.png")
```

You can also save your plot and then call `ggsave()`. Notice we can change the dimensions and units. 

```{r}
# save plot
p <- ggplot(homes) +
  aes(x = reorder(esdistrict, totalvalue, median), y = totalvalue) +
  geom_boxplot() +
  labs(x = "Elementary School District", y = "Total Home Value") +
  scale_y_continuous(labels = dollar) +
  coord_flip() +
  theme_minimal()
p
ggsave(filename = "es_totalvalue_boxplot.jpg", plot = p, 
       width = 20, height = 15, units = "cm")
```


When working in an R script (as opposed to an Rmd file), plots appear in the Plot frame within RStudio which provides a convenient Export button to interactively save plots.


## We're done!

We just covered the basics of ggplot2 today. The way to get better at using ggplot2 is to keep using it and Googling/ask AI for help when you get stuck.

If you would like to talk more about ggplot2 or anything else statistics related, I would love to hear from you: `clayford@virginia.edu` or `statlab@virginia.edu`


## References 

- "GIS Data | Albemarle County, VA." (2025) Albemarle.org. <https://www.albemarle.org/government/information-technology/geographic-information-system-gis-mapping/gis-data>.

- Wickham, H. (2016), _ggplot2: Elegant Graphics for Data Analysis_ (3rd ed), Springer. [https://ggplot2-book.org/](https://ggplot2-book.org/) 

- Wickham, H. and Grolemund G. (2017), _R for Data Science_. O'Reilly. [http://r4ds.had.co.nz/](http://r4ds.had.co.nz/)

- ggplot2 cheat sheet: 
<https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf>


## Appendix

Additional material cut for time.

### Move the legend

To move the legend, use `+ theme(legend.position)` where the `legend position` argument take the values "none", "bottom", "left", or "top".

Example:

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = hsdistrict) +
  geom_point() +
  theme(legend.position = c(12000, 1e6))
```

To place a legend inside a plot, set `legend.position = "inside"` and then set `legend.position.inside` to relative coordinates between 0 and 1.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = hsdistrict) +
  geom_point() +
  theme(legend.position = "inside", 
        legend.position.inside = c(0.85, 0.15))
```

### Rotate axis labels

Recall the boxplot of totalvalue by esdistrict:

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_boxplot()
```

We can rotate the axis labels 45 degrees to make them legible by specifying `theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))`. The angle argument can take values ranging from [0, 360]. The vjust and hjust arguments can take values ranging from [0,1]. The specify vertical and horizontal justification. Getting labels rotated and looking the way you want often takes some experimenting.

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
```



### Order bar charts by counts

If we want order the x-axis in ascending order by counts, we need to use the `after_stat()` function with an argument of `count` in a call to `aes()` within the `geom_bar()` function, and then use `sort()`.

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar(aes(y = sort(after_stat(count)))) +
  ylab("count")
```

To sort in descending order:

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar(aes(y = sort(after_stat(count), decreasing = TRUE))) +
  ylab("count")
```

This is made a little easier by working directly with data frames of counts.

```{r}
count_df <- as.data.frame(xtabs(~ condition, data = homes))
count_df
```

Now, instead of using `geom_bar()`, use `geom_col()`. Notice we need to define a `y` aesthetic for the counts. ("Freq" is the default name given to the counts by the `as.data.frame()` function.) Finally use the `reorder()` function to reorder condition by Freq.

```{r}
ggplot(count_df) +
  aes(x = reorder(condition, Freq), y = Freq) +
  geom_col() +
  xlab("condition")

```


To reorder in descending order, set `decreasing = TRUE` in the `reorder()` function.

```{r}
ggplot(count_df) +
  aes(x = reorder(condition, Freq, decreasing = TRUE), y = Freq) +
  geom_col() +
  xlab("condition")
```


### Plotting with two data frames 

ggplot allows us to use multiple data frames when creating a plot. For example, consider the following plot totalvalue vs YearBuilt for the years 1950 - 2025 for homes valued at $1,000,000 or less. Notice we are zoomed in.

```{r}
ggplot(homes) +
  aes(x=yearbuilt, y=totalvalue) + 
  geom_jitter(width = 0.2, height = 0, shape=".") +
  coord_cartesian(xlim=c(1950,2016), ylim=c(0,1e6))

```

Let's say we wanted to plot on top of that graph the median totalvalue by year. First we need to calculate the median by year. Here's one way to do it:


```{r}
homeValues <- aggregate(totalvalue ~ yearbuilt, data = homes, median)
tail(homeValues, n = 10)
```

Each `geom_xxx()` function allows you to supply them with a separate data frame and mappings. Below notice the `geom_line()` function references the homeValues data frame and provide a different mapping to the y aesthetic. Also notice we have to swap the order of data and aesthesics.

```{r}
ggplot(homes) + 
  aes(x=yearbuilt, y=totalvalue) + 
  geom_jitter(width = 0.2, height = 0, shape=".") +
  # Notice the next geom uses a new data frame
  geom_line(aes(y = totalvalue), homeValues, 
            color = "red") +
  coord_cartesian(xlim=c(1950,2016), ylim=c(0,1e6)) 

```

Another example: Boxplot of totalvalues by middle school district, zoomed in on homes $1,000,000 or less

```{r}
ggplot(homes) +
  aes(x = msdistrict, y = totalvalue) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0,1e6))

```


Let's say we wanted to label the medians with the actual amount.

First we calculate it. Notice we create two versions, one numeric and one formatted in dollar amounts using the dollar function from the scales package.

```{r}
homeMedians <- aggregate(totalvalue ~ msdistrict, data = homes, median)
# add column with median value formatted as dollar amount
homeMedians$medianValueD <- scales::dollar(homeMedians$totalvalue)
homeMedians
```

Now we use the geom_text() function to add the amount to the boxplot. Notice  we use alpha = 1/4 to lighten the boxplot, that the label aesthetic says which data will be mapped to the text label, and that we add 25,000 to the  positioning so the label is above the median line. (that took some trial and error)

```{r}
ggplot(homes) +
  aes(x = msdistrict, y = totalvalue) +
  geom_boxplot(alpha = 1/4) +
  # Notice the next geom uses a new data frame
  geom_text(aes(label = medianValueD, y = totalvalue + 25000), homeMedians) +
  coord_cartesian(ylim = c(0,1e6))

```

### plotting functions in ggplot2

At the opening of the workshop I drew a parabola with base R code:

```{r}
x <- seq(-1, 1, by = 0.01)
y <- x^2
plot(x, y, type = "l")
```

Here's how to do the same with {ggplot2}. Notice we need to define the limits of the x-axis first. Then we use `geom_function()` to define the function.

```{r}
ggplot() +
  xlim(c(-1,1)) +
  geom_function(fun = function(x)x^2)
```

Here's how to plot a standard normal curve. We use the `dnorm()` function without the parentheses.

```{r}
ggplot() +
  xlim(c(-3,3)) +
  geom_function(fun = dnorm)

```

Here's how to plot a sin function.

```{r}
ggplot() +
  xlim(c(-3*pi,3*pi)) +
  geom_function(fun = sin)

```


### Adding error bars to plots

Below we model mpg as a function of cyl using the mtcars dataset that comes with R. This is basically a one-way ANOVA. Once we fit our model, we use the {emmeans} package to estimate mean mpg values per cylinder level. Notice it returns lower and upper values of a 95% confidence interval.

```{r}
mtcars$cyl <- factor(mtcars$cyl)
m <- lm(mpg ~ cyl, data = mtcars)
library(emmeans)
emmeans(m, specs = "cyl")
```

We can convert that output to a data frame using the `as.data.frame()` function.

```{r}
ci_df <- as.data.frame(emmeans(m, specs = "cyl"))
ci_df
```

Now we can create the plot using the `geom_point()` and `geom_errorbar()` functions. Notice we need to specify `ymin` and `ymax` aesthetics for the `geom_errorbar()` function.

```{r}
ggplot(ci_df) +
  aes(x = cyl, y = emmean, ymin = lower.CL, ymax = upper.CL) +
  geom_point() +
  geom_errorbar(width = 0.1) +
  labs(y = "Estimated MPG", x = "Cylinder", 
       title = "95% Confidence Intervals of Estimated MPG by Cylinder")
```

### Change default color palette

If you want to change the color palette so you don't have to repeatedly use `scale_color_manual()` or `scale_fill_brewer()` in your code, you can set ggplot2 options using the `options()` function. These steps come from the following Stack Overflow answer: <https://stackoverflow.com/a/68205270/2765195>

Let's say you want to use the "Set2" palette from the {RColorBrewer} package any time you map a categorical variable to a color or fill scale.

```{r}
opts <- options()  # save old options

# definie new scale functions with your palette of choice
scale_colour_brewer2 <- function(...) {
  scale_colour_brewer(..., palette = "Set2")
}

scale_fill_brewer2 <- function(...) {
  scale_fill_brewer(..., palette = "Set2")
}

# set the options
options(
  ggplot2.discrete.colour = scale_colour_brewer2,
  ggplot2.discrete.fill = scale_fill_brewer2
)
```

Now the "Set2" color palette is the default palette for the color scale.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = cooling) +
  geom_point() 

```

And it's the default palette for the fill scale. 

```{r}
ggplot(homes) +
  aes(x = condition, fill = hsdistrict) +
  geom_bar(position = "fill") 
```

Run the following line to restore the original default options. Or you can simply restart R.

```{r}
options(opts)
```

