---
title: "Data Visualization in R"
author: "Clay Ford, Statistical Research Consultant, UVA Library"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Quick Intro to R Notebooks and R Markdown

This is an R Markdown Notebook. It combines markdown, a plain text formatting syntax, and R code. When you execute R code within the notebook, the output appears beneath the code. 

This file was created in RStudio by going to File...New File...R Notebook.

R code needs to be in "chunks" for it to work. Below is an example of an R code chunk. It makes a parabola using base R graphics (not ggplot2).

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter* (Win/Linux) or *Cmd+Shift+Return* (Mac). 

```{r}
x <- seq(-1, 1, by = 0.01)
y <- x^2
plot(x, y, type = "l")
```

To hide the output, click the Expand/Collapse output button. To clear results (or an error), click the "x". 

You can also press *Ctrl+Enter* (Win/Linux) or *Cmd+Return* (Mac) to run one line of code at a time (instead of the entire chunk).

Add a new R code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac).  

## Exercise 0 (code along)

Insert a new R code chunk below and type/run the code: 22/7



## Import data

Let's import some data which we'll visualize today using ggplot2. The data we'll work with is Albemarle County real estate data which was downloaded from the Office of Geographic Data Services. We'll use a random sample of the data.

```{r}
URL <- 'https://raw.githubusercontent.com/clayford/dataviz_with_ggplot2/master/alb_homes.csv'
homes <- read.csv(file = URL)
```

Let's look at the first few rows:

```{r}
head(homes)
```

Variable name definitions:

- *yearbuilt*: year house was built
- *finsqft*: size of house in number square feet
- *cooling*: 'Central Air' versus 'No Central Air'
- *bedroom*: number of bedrooms
- *fullbath*: number of full bathrooms (toilet, sink and bath)
- *halfbath*: number of half bathrooms (toilet and sink only)
- *lotsize*: size of land on which home is located, in acres
- *totalvalue*: total assessed value of home and property
- *esdistrict*: the elementary school the home feeds into
- *msdistrict*: the middle school the home feeds into
- *hsdistrict*: the high school the home feeds into
- *censustract*: the census tract the home is located in
- *age*: of the house in years as of 2018
- *condition*: assessed condition of home (Substandard, Poor, Fair, Average, Good, Excellent)
- *fp*: indicator if house has fireplace (0=no, 1=yes)


## Intro to ggplot2

The ggplot2 package implements *The Grammar of Graphics* as defined in the book of the same name by Leland Wilkinson.

It requires your data be in a data frame.

Let's do a quick example. How does the total value of a home relate to its size in finished square feet? Plot totalvalue versus finsqft to create a scatter plot. 

First we need to load the ggplot2 package. Only do this once per R session.

```{r}
# install.packages('ggplot2')
library(ggplot2)
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point()
```

What just happened?

- The `ggplot` function creates a plotting area for a data frame
- The `aes` function maps variables in the data frame to aesthetic properties of geometric shapes
- The `geom_point` function specifies we want to use points
- combine these functions with `+`

Don't forget the parentheses on `geom_point()`! 

*Most of what we do with ggplot2 takes this general form*

```
ggplot(data) +
  aes() +
  geom_something()
```

Obviously it can get much more complicated. But for exploratory visualization this often sufficient.


## Exercise 1 (code along)

Do younger (more recently built) homes have a tendency to be bigger than older homes? Create a scatterplot of finsqft (y-axis) versus age (x-axis). Reminder: Insert a new code chunk by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac). Try entering alpha = 1/5 in `geom_point()`. What does it do?




## More aesthetics and facets

Points not only have positions in a plane, but can be different sizes and have different shapes and colors. We can map variables in our data frame to these aesthetics. 

How does the relationship between totalvalue and finsqft differ between homes with and without central air?

Plot totalvalue versus finsqft to create a scatter plot, but also color the points according to whether or not the homes have central air (cooling). Notice we add `color = cooling` to the `aes()` function below: 

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = cooling) +
  geom_point() 
```

We can also map the size of the points to a variable, such as lotsize. Notice I have mapped the shape of the point to cooling using the argument `size = lotsize`.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, shape = cooling, size = lotsize) +
  geom_point()

```

Notice the overplotting. Perhaps we prefer to see separate scatterplots for each cooling category. We can achieve that with *facets*. The function to use is `facet_wrap()` with the variable that we want to facet on. In this case it's cooling. Notice we precede the variable with a tilde ~. We can read that as "facet by cooling"  

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, size = lotsize) +
  geom_point(alpha = 1/2) +
  facet_wrap(~ cooling)

```

We can also have multiple geoms in a plot. Let's say we want to add a *smooth trend line* that summarizes the association between finsqft and totalvalue. We can add `geom_smooth()` to our code:

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point(alpha = 1/2) +
  geom_smooth() +
  facet_wrap(~ cooling) 

```


## Exercise 2

How is totalvalue related to lotsize within each of the three high school districts (hsdistrict)?

Create a scatterplot of totalvalue (y-axis) vs lotsize (x-axis), faceted by hsdistrict. Add smooth trend lines to summarize the relationship.




### Zooming in on data

Albemarle county has homes with very large totalvalue and lotsize values. Including them in our plots means we "zoom out" to accommodate them which means we're "far away" from the majority of homes.  

To zoom in on data with ggplot2, we use the `coord_cartesian()` function and define the limits to view with the `xlim` and `ylim` arguments. Below we zoom in on the y-axis to see homes that range in value from from 0 to $2,000,000 and on the x-axis to see homes that range in finsqft from 0 to 4000. Notice we define that lower and upper limits as a vector using the `c()` function.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point() +
  geom_smooth() + 
  coord_cartesian(xlim = c(0, 4000), ylim = c(0, 2e6))

```


The numbers are a little hard to read on the y-axis. We can format them as dollar amounts using the labels argument in the `scale_y_continuous()` function. To do this we need to load the scales package, which is installed when you install the ggplot2 package. 

```{r}
library(scales) # for the dollar function
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(xlim = c(0, 4000), ylim = c(0, 2e6)) + 
  scale_y_continuous(labels=dollar) 

```



## Exercise 3

Modify your code from Exercise 2 to zoom in on the y-axis to see home prices ranging from 0 to $1,000,000, and zoom in on the x-axis to view homes on lots ranging in size from 0 to 5 acres. Also use the dollar function to format the home prices on the y-axis.

```{r}
ggplot(homes) +
  aes(x = lotsize, y = totalvalue) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ hsdistrict)

```

## Visualizing counts

Recall the condition variable. We can quickly get counts of each condition using the `table()` function:

```{r}
table(homes$condition)
```

One way to visualize counts is with a bar plot. We can create a bar plot in ggplot2 with `geom_bar()`. Notice it automatically generates the counts for us.

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar()
```

If we want the x-axis ordered according to condition, it's best to define that ordering in the data frame. We can use the `factor` function to achieve this.

```{r}
homes$condition <- factor(homes$condition, 
                          levels = c('Substandard', 'Poor', 'Fair', 
                                     'Average', 'Good', 'Excellent'))
```

Now when we generate the bar plot, the x-axis in the order as defined in the data frame.

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar()
```

One of the aesthetics of bars is their fill color. We can map a variable from our data frame to the fill aesthetic. For example, let's see counts of conditions for homes with and without central air (cooling). Notice we add `fill = cooling` to the `aes()` function.

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar()

```

The default result is to stack the bars. We can set them side-by-side setting the position argument to "dodge".

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar(position = "dodge")

```


The large number of Average homes with central air makes it difficult to see the counts for the other categories. An alternative approach is to view the *proportion* of homes with and without central air within each condition. We can do this by setting the position argument to "fill".

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar(position = "fill")

```

The y-axis now represents proportion. 100% of substandard homes have no central air.

## Exercise 4

Given a home's condition, is it more likely to be in a particular high school district? For example, given all the homes in Excellent condition, is there a higher proportion of them in one of the three high school districts?

Create a bar plot of condition (condition on x-axis), filled by hsdistrict. Set position = 'fill' so we can visualize within each condition the proportion belonging to a high school district.



## Continuous vs groups

How does the totalvalue of homes differ between elementary school districts? One way to visualize this is with a boxplot. We can do this with `geom_boxplot`.


```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_boxplot()
```

The x-axis is hard to read. This happens sometimes. One solution is to simply rotate the plot. We can do that with the `coord_flip()` function. The `coord_flip()` function also allows us to zoom in on a plot with the xlim and ylim arguments. Let's zoom in on the y-axis and look at homes less than 2 million in totalvalue.

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_boxplot() +
  coord_flip(ylim = c(0,2e6))

```


- The width of the box (along the y axis) represents the middle 50% of the data. It is called the Interquartile Range (IQR).
- The line in the box is the median. 
- The top of the box is the 75th percentile. 
- The bottom of the box is the 25th percentile. 
- The "whiskers" extending out of the box are to the highest and lowest values not to exceed 1.5 * IQR
- The dots are considered "outliers" relative to the rest of the data within each elementary school district.


A similar plot is the violin plot, which allows us to compare the distributions between categories using smooth density plots. Use `geom_violin()`

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_violin() +
  coord_flip(ylim = c(0,1e6))
```

Boxplots and violin plots obscure the total number within in each group. That's not necessarily a bad thing. Too much information in a plot can be overwhelming. However ggplot allows us to combine geoms, so we could combine a 1-dimensional scatterplot (sometimes called a "stripchart") with a boxplot or violin plot. The `position_jitter` functions adds a little random noise to data to help spread them out. 

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = totalvalue) +
  geom_point(position = position_jitter(width = 0.2, height = 0)) +
  geom_violin(alpha = 1/2) +
  coord_flip(ylim = c(0,1e6))

```


## Exercise 5

How are the ages of homes distributed between the elementary school districts? Are there some elementary schools that serve mostly newer homes? Create a boxplot of age of home by esdistrict. 



## Line plots over time

How has the median size (finsqft) of houses changed over time (yearbuilt)? It seems like homes have been getting bigger. Can we visualize how the median finsqft has changed over the years?

To do this we need to get the median finsqft per year. The result needs to be a data frame. There are many ways to do this in R. Here is one way.

```{r}
homes_median_sf <- aggregate(finsqft ~ yearbuilt, data = homes, median)
tail(homes_median_sf, n = 10)
```

Let's plot finsqft vs yearbuilt using dots and lines.

```{r}
ggplot(homes_median_sf) +
  aes(x = yearbuilt, y = finsqft) +
  geom_point() +
  geom_line()
```

There's a lot of variability in median home size in finsqft over time! But it tends to settle down around 1950 and then slowly increase. 


How has lotsize changed over time? Let's plot median lotsize by yearbuilt versus yearbuilt. Again we need to make a new data frame that contains yearbuilt and median lotsize per year.

```{r}
homes_median_lotsize <- aggregate(lotsize ~ yearbuilt, data = homes, median)

ggplot(homes_median_lotsize) +
  aes(x = yearbuilt, y = lotsize) +
  geom_line() 
```

After 1950 it looks most new homes are built on small lots.

We can zoom in and get more detail on the last 90 years. Notice we use the `labs()` function to add a title and customize the axis labels.

```{r}
ggplot(homes_median_lotsize) +
  aes(x = yearbuilt, y = lotsize) +
  geom_line() +
  coord_cartesian(xlim = c(1930,2020), ylim = c(0,30)) +
  labs(title = "Median Lot Size by Year Built", 
       x = "Year Built", 
       y = "Median Lot Size")

```



## Exercise 6

How has the mean number of fullbaths in a home changed over time (yearbuilt), since 1950?

The following code creates a data frame of mean fullbath by yearbuilt for 1950 to present.

```{r}
homes_mean_fb <- aggregate(fullbath ~ yearbuilt, 
                           data = homes, mean, 
                           subset = yearbuilt >= 1950)
head(homes_mean_fb)
```

Create a line plot showing how the mean number of fullbaths has changed over time (yearbuilt). Try adding a smooth trend line using `geom_smooth()`.




## Wrap up

We just covered the basics of ggplot2 today. The way to get better at using ggplot2 is to keep using it and Googling for help when you get stuck. (Seriously)

Two more topics worth mentioning:

1. Changing the appearance of plots
2. Saving plots as jpg, png, tif, etc.

The appearance of plots are dictated by themes. Some different themes are included with ggplot2. For example, `theme_minimal()`. Simply tack it on to the end of a plot.

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = lotsize) +
  geom_point(position = position_jitter(width = 0.1, height = 0)) +
  coord_flip(ylim = c(0,50)) +
  theme_minimal()
```

There are whole R packages that provided nothing but more themes for ggplot2. For example, see the `ggthemes` package.

Saving plots can be done with the `ggsave()` function. By default, it will save the last plot you created as a 7 in. x 7 in. image.

```{r}
ggsave("es_lotsize.png")
```

`ggsave()` can also be tacked on to the end of ggplot code.

```{r}
ggplot(homes) +
  aes(x = esdistrict, y = lotsize) +
  geom_point(position = position_jitter(width = 0.1, height = 0)) +
  coord_flip(ylim = c(0,50)) +
  theme_minimal() +
  ggsave("es_lotsize.jpeg", width = 20, height = 20, units = "cm")

```

When working in an R script (as opposed to an Rmd file), plots appear in the Plot frame within RStudio which provides a convenient Export button to interactively save plots.


## We're done!

If you would like to talk more about ggplot2 or anything else statistics related, I would love to hear from you: `clayford@virginia.edu` or `statlab@virginia.edu`

Want more ggplot2? Google `getting started with ggplot2` or `ggplot2 tutorial`. Also see the numerous examples provided with the ggplot2 documentation.


## References 

* Wickham, H. (2016), _ggplot2: Elegant Graphics for Data Analysis_ (2nd ed), Springer. [https://ggplot2-book.org/](https://ggplot2-book.org/) 

* Wickham, H. and Grolemund G. (2017), _R for Data Science_. O'Reilly. [http://r4ds.had.co.nz/](http://r4ds.had.co.nz/)

**ggplot2 cheat sheet**    
https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf


