---
title: "Data Visualization in R with ggplot2"
author: "Clay Ford, Statistical Research Consultant, UVA Library"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Quick Intro to R Notebooks and R Markdown

This is an R Markdown Notebook. It combines markdown, a plain text formatting syntax, and R code. When you execute R code within the notebook, the output appears beneath the code. 

This file was created in RStudio by going to File...New File...R Notebook.

R code needs to be in "chunks" for it to work. Below is an example of an R code chunk.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter* (Win/Linux) or *Cmd+Shift+Return* (Mac). 

```{r}
summary(cars)
plot(cars$speed, cars$dist)
```

To hide the output, click the Expand/Collapse output button. To clear results (or an error), click the "x". 

You can also press *Ctrl+Enter* (Win/Linux) or *Cmd+Return* (Mac) to run one line of code at a time (instead of the entire chunk).

Add a new R code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac).  

## Exercise 0 (code along)

Insert a new R code chunk below and type/run the code: 22/7



## Import data

Let's import some data which we'll visualize today using ggplot2. The data we'll work with is Albemarle County real estate data which was downloaded from the Office of Geographic Data Services. We'll use a random sample of the data.

```{r}
URL <- 'https://raw.githubusercontent.com/clayford/dataviz_with_ggplot2/master/alb_homes.csv'
homes <- read.csv(file = URL)
```

Let's look at the first few rows:

```{r}
head(homes)
```

Variable name definitions:

- *yearbuilt*: year house was built
- *finsqft*: size of house in number square feet
- *cooling*: 'Central Air' versus 'No Central Air'
- *bedroom*: number of bedrooms
- *fullbath*: number of full bathrooms (toilet, sink and bath)
- *halfbath*: number of half bathrooms (toilet and sink only)
- *lotsize*: size of land on which home is located, in acres
- *totalvalue*: total assessed value of home and property
- *esdistrict*: the elementary school the home feeds into
- *msdistrict*: the middle school the home feeds into
- *hsdistrict*: the high school the home feeds into
- *censustract*: the census tract the home is located in
- *age*: of the house in years as of 2018
- *condition*: assessed condition of home (Substandard, Poor, Fair, Average, Good, Excellent)
- *fp*: indicator if house has fireplace (0=no, 1=yes)


## Intro to ggplot2

The ggplot2 package implements *The Grammar of Graphics* as defined in the book of the same name by Leland Wilkinson.

It requires your data be in a data frame or tibble.

Let's do a quick example: Plot totalvalue versus finsqft to create a scatter plot. How does the total value of a home relate to its size in finished square feet?

First we need to load the ggplot2 package. Only do this once per R session.

```{r}
# install.packages('ggplot2')
library(ggplot2)
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point()
```

Don't forget the parentheses on `geom_point()`! 

What just happened?

- The `ggplot` function creates a plotting area for a data frame
- The `aes` function map variables in the data frame to aesthetic properties of geometric shapes
- The `geom_point` function specifies we want to use points
- combine these functions with `+`

NOTE: We can also nest the `aes` function in the `ggplot` function:

```{r}
ggplot(homes, aes(x = finsqft, y = totalvalue)) +
  geom_point()
```



## Exercise 1 (code along)

Create a scatterplot of finsqft (y-axis) versus age (x-axis). Do younger (more recently built) houses have a tendency to be bigger than older houses? Reminder: Insert a new code chunk by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac). Try entering alpha = 1/5 in `geom_point()`. What does it do?


## More aesthetics

Points not only have positions in a plane, but can be different sizes and have different shapes and colors. We can map variables in our data frame to these aesthetics. 

Plot totalvalue versus finsqft to create a scatter plot, but also color the points according to whether or not the homes have central air (cooling). How does the relationship between totalvalue and finsqft differ between homes with and without central air?

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, color = cooling) +
  geom_point() 
```

We can also map the size of the points to a variable, such as lotsize.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, shape = cooling, size = lotsize) +
  geom_point()

```

Notice the overplotting. Perhaps we prefer to see separate scatterplots for each cooling category. We can achieve that with *facets*. The function to use is `facet_wrap()` with the variable that we want to facet on. In this case it's cooling. Notice we precede the variable with a tilde ~. We can read that as "facet by cooling"  

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue, size = lotsize) +
  geom_point(alpha = 1/2) +
  facet_wrap(~ cooling)

```

We can also have multiple geoms in a plot. Let's say we want to add a smooth trend line that summarizes the relationship between finsqft and totalvalue:

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point(alpha = 1/2) +
  geom_smooth() +
  facet_wrap(~ cooling)

```

## Exercise 2

Create a scatterplot of totalvalue (y-axis) vs lotsize (x-axis), faceted by hsdistrict. Add smooth trend lines to summarize the relationship.



### Zooming in on data

Albemarle county has homes with very large totalvalue and lotsize values. Including them in our plots means we "zoom out" to accommodate them which means we're "far away" from the majority of homes.  

To zoom in on data with ggplot2, we use the `coord_cartesion()` function and define the limits to view with the `xlim` and `ylim` arguments. Below we zoom in on the y-axis only.

```{r}
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point() +
  geom_smooth() + 
  coord_cartesian(ylim = c(0, 2e6)) # from 0 to $2,000,000

```


The numbers are a little hard to read on the y-axis. We can format them as dollar amounts using the labels argument in the `scale_y_continuous()` function. To do this we need to load the scales package, which is installed when you install the ggplot2 package. 

```{r}
library(scales) # for the dollar function
ggplot(homes) +
  aes(x = finsqft, y = totalvalue) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(ylim = c(0, 2e6)) + 
  scale_y_continuous(labels=dollar) 

```



## Exercise 3

Modify your code from Exercise 2 to zoom in on the y-axis to see home prices ranging from 0 to $1,000,000, and zoom in on the x-axis to view homes on lots ranging in size from 0 to 5 acres. Also use the dollar function to format the home prices on the y-axis.

```{r}
ggplot(homes) +
  aes(x = lotsize, y = totalvalue) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ hsdistrict) 

```

## Visualizing counts

Recall the condition variable. We can quickly get counts of each condition using the table function:

```{r}
table(homes$condition)
```

One way to visualize counts is with a bar plot. We can create a bar plot in ggplot2 with `geom_bar()`. Notice it automatically generates the counts for us.

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar()
```

If we want the x-axis ordered according to condition, it's best to define that ordering in the data frame. We can use the `factor` function to achieve this.

```{r}
homes$condition <- factor(homes$condition, 
                          levels = c('Substandard', 'Poor', 'Fair', 
                                     'Average', 'Good', 'Excellent'))
```

Now when we generate the bar plot, the x-axis in the order as defined in the data frame.

```{r}
ggplot(homes) +
  aes(x = condition) +
  geom_bar()
```

One of the aesthetics of bars is their fill color. We can map a variable from our data frame to the fill aesthetic. For example, let's see counts of conditions for homes with and without central air (cooling).

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar()

```

The default result is to stack the bars. We can set them side-by-side setting the position argument to "dodge".

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar(position = "dodge")

```


The large number of Average homes makes it difficult to see the counts for the other categories. An alternative approach is to view the proportion of homes with and without central air within each condition. We can do this by setting the position argument to "fill".

```{r}
ggplot(homes) +
  aes(x = condition, fill = cooling) +
  geom_bar(position = "fill")

```

The y-axis now represents proportion. 100% of substandard homes have no central air.

## Exercise 4

Create a bar plot of condition (condition on x-axis), filled by hsdistrict. Set position = 'fill' so we visualize within each condition the proportion belonging to a high school district.




## Continuous vs group (boxplots, violin plot, etc)

...



## Line plots over time

...



## References for going further

- ggplot2 cheat sheet: https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf 

- online version of work-in-progress 3rd edition of *ggplot2: elegant graphics for data analysis*: https://ggplot2-book.org/

- Hundreds of charts with reproducible code: https://www.r-graph-gallery.com/


